---
title: "Tau Mutant GeoMx Analysis"
author:
  - name: "Emir Turkes and Naoto Watamura, UK DRI at UCL"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
link-citations: true
output:
  html_document:
    code_folding: show
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(
    inputFile,
    encoding = encoding,
    output_file = file.path("..", "results", "tau_mutant_geomx_analysis.html")
  )})
---

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 25px; text-indent: 0;
}
div.main-container {max-width: none; width: 100%;}
</style>

```{r}
# Copyright 2025 Emir Turkes, Naoto Watamura, UK DRI at UCL
#
# Licensed under the Apache License, Version 2.0 (the "License");
# you may not use this file except in compliance with the License.
# You may obtain a copy of the License at
#
#     http://www.apache.org/licenses/LICENSE-2.0
#
# Unless required by applicable law or agreed to in writing, software
# distributed under the License is distributed on an "AS IS" BASIS,
# WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
# See the License for the specific language governing permissions and
# limitations under the License.

# Load packages.
# --------------
library(conflicted)
packages <- c("GeomxTools", "Seurat")
invisible(
  suppressPackageStartupMessages(
    lapply(packages, FUN = library, character.only = TRUE)
  )
)
# --------------

# Add paths.
# ----------
data_dir <- file.path("..", "data")
cache_dir <- file.path("..", "cache")
if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}
# ----------

# Global settings.
# ----------------
knitr::opts_chunk$set(fig.width = 10, fig.height = 7, dpi = 96)
# ----------------
```

# Data Prep

```{r}
# Paths to data.
# --------------
dcc <- dir(
  file.path(data_dir, "external", "tau-mutant-geomx", "GeoMxPipeline"),
  pattern = ".dcc$", full.names = TRUE
)
pkc <- file.path(
  data_dir, "external", "tau-mutant-geomx", "Mm_R_NGS_WTA_v1.0.pkc"
)
anno <- file.path(
  data_dir, "external", "tau-mutant-geomx",
  "dufflab_20241126T1146", "AnalysisSampleSheet.xlsx"
)
# --------------

# Create NanoStringGeoMxSet.
# --------------------------
rds <- file.path(cache_dir, "data.rds") # Cache object due to slow execution.
if (file.exists(rds)) {
  data <- readRDS(rds)
} else {
  data <- suppressWarnings(
    readNanoStringGeoMxSet(
      dcc, pkc, anno,
      phenoDataSheet = "Sheet1",
      protocolDataColNames = c("genotype", "bio_rep", "tech_rep", "slide_rep")
    )
  )
  saveRDS(data, file = rds)
}
# --------------------------

data
```

# Quality Control

```{r}
# ROI level QC.
# -------------
data <- shiftCountsOne(data, useDALogic = TRUE)
data <- setSegmentQCFlags(data, qcCutoffs = list(minSegmentReads = 30000))

QC_results <- protocolData(data)$QCFlags
QC_summary <- data.frame(
  Pass = colSums(!QC_results[, colnames(QC_results)]),
  Warning = colSums(QC_results[, colnames(QC_results)])
)
QC_results$Status <- apply(
  QC_results, MARGIN = 1L, FUN = function(x) {
    ifelse(sum(x) == 0L, yes = "PASS", no = "WARNING")
  }
)
QC_summary["TOTAL FLAGS", ] <- c(
  sum(QC_results[, "Status"] == "PASS"),
  sum(QC_results[, "Status"] == "WARNING")
)

data <- data[, QC_results$LowReads == "FALSE"] # Only filter on library size.
# -------------

QC_summary

# Probe level QC.
# ---------------
data <- setBioProbeQCFlags(data)

QC_results <- fData(data)$QCFlags
QC_summary <- data.frame(
  Passed = sum(rowSums(QC_results[, -1]) == 0),
  Global = sum(QC_results$GlobalGrubbsOutlier),
  Local = sum(
    rowSums(QC_results[, -2:-1]) > 0 & !QC_results$GlobalGrubbsOutlier
  )
)
# ---------------

QC_summary # No removal needed.

data <- aggregateCounts(data) # Turn probe names into gene symbols.
data.frame(assayData(data)$exprs[1:5, 1:2])

data <- normalize( # Normalize data.
  data, norm_method = "quant", desiredQuantile = 0.75, toElt = "q_norm"
)
data.frame(assayData(data)$q_norm[1:5, 1:2])
```

# Create Seurat Object

```{r}
data <- as.Seurat(data, normData = "q_norm")
data <- RenameAssays(
  data, assay.name = "GeoMx", new.assay.name = "RNA", verbose = FALSE
)
data$genotype <- factor( # Rearrange order of labels.
  data$genotype, levels = c("MAPTKI", "P301S", "NLGF_MAPTKI", "NLGF_P301S")
)
data
```

# Dimensionality Reduction

```{r}
data <- SCTransform(data, verbose = FALSE)
data <- RunPCA(data, npcs = 30, verbose = FALSE)
data <- FindNeighbors(data, verbose = FALSE)
data <- FindClusters(data, verbose = FALSE)
data <- RunUMAP(data, dims = 1:30, verbose = FALSE)

DimPlot(data, group.by = "genotype")
```

# Save Seurat Object

```{r}
rds <- file.path(cache_dir, "seurat.rds")
if (!file.exists(rds)) {
  saveRDS(data, file = rds)
}
```

# References

This is the concluding section of the document, where we output the `sessionInfo`, and create a bibliography for works cited.

```{r}
sessionInfo()
```
